---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# poEMirt

<!-- badges: start -->
[![R-CMD-check](https://github.com/vkyo23/poEMirt/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/vkyo23/poEMirt/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`poEMirt` is an `R` package that implements fast EM item response theory for public opinion data analysis to estimate cross-national latent public opinion over years.
This package utilizes the Polya-Gamma data augmentation scheme (Polson, Scott & Windle 2013) and exact EM algorithm (Goplerud 2019) for fast computation. 
This package incorporates direct C++ programming and does not depend on ready-made other statistical modeling languages, thus the computation is more efficient.
It can be applied to not only binomial and multinomial outcomes but also bernoulli and categorial ones. 
Users can estimate statistical uncertainty by parametric bootstrap and Gibbs Sampling.

## Installation

You can install the development version of `poEMirt` from [GitHub](https://github.com) with: 

```{r, eval=FALSE}
remotes::install_github("vkyo23/poEMirt")
```

## Usage

### Multinomial outcome

```{r}
library(poEMirt)
data("sim_data_dynamic")

head(sim_data_dynamic)
tail(sim_data_dynamic)

# Convert into poEMirt-readable data
data <- read_poEMirt(
  data = sim_data_dynamic,
  responses = paste0('y', 1:5),
  i = "i",
  j = "j",
  t = "t"
)
summary(data)

# Fit the model
fit <- poEMirt(
  data = data,
  model = "dynamic",
  control = list(
    verbose = 10
  )
)

# Summarize the result
summary(fit, parameter = "theta")
```

### Estimation of statistical uncertainty

To estimate statistical uncertainty (e.g., standard deviation), you can use `poEMirt_uncertainty()`. In this function, you can choose *parametric bootstrap* (`"bootstrap"`) or *Gibbs sampling* (`"gibbs"`).

#### Bootstrap

```{r}
fit_boot <- poEMirt_uncertainty(
  fit = fit,
  method = "bootstrap",
  seed = 1,
  iter = 100,
  control = list(
    verbose = 10
  )
)

summary(fit_boot, parameter = "theta", ci = 0.95)
```

```{r, include=FALSE}
rm(list = "fit_boot")
gc(reset = TRUE); gc(reset = TRUE)
```


#### Gibbs sampling

Using EM estimates as starting values of Gibbs sampler.

```{r}
fit_gibbs <- poEMirt_uncertainty(
  fit = fit,
  method = "gibbs",
  seed = 1,
  iter = 500,
  control = list(
    verbose = 50,
    warmup = 100,
    thin = 5
  )
)

summary(fit_gibbs, parameter = "theta", ci = 0.95)
```

```{r, include=FALSE}
rm(list = "fit_gibbs")
gc(reset = TRUE); gc(reset = TRUE)
```

### Binary, categorical outcome

Notably, `poEMirt` can also be applied to roll-call vote, individual-level survey, and exam data analysis where outcomes are binary or categorical since these are the special case of `poEMirt` where n = 1 for each observation.

#### Binary

Using the roll-call vote data of U.S. Supreme Court `Rehnquist` from [`{MCMCpack}`](https://github.com/cran/MCMCpack) by Martin, Quinn & Park (2011). The outcome of interest is 1 if a judge votes yea and 0 if nay.

```{r}
library(tidyr)
library(dplyr)
data(Rehnquist, package = "MCMCpack")

df <- Rehnquist |> 
  mutate(rcid = row_number()) |> 
  select(-term) |> 
  mutate(
    time = factor(time) |> 
      as.integer()
  ) |> 
  pivot_longer(
    cols = -c(time, rcid),
    names_to = "judge",
    values_to = "vote"
  ) 
head(df)

#For binary data, users must create the columns of `outcome = 1` and `outcome = 0` data respectively. 
df <- df |> 
  mutate(
    y1 = 1L * (vote == 1),
    y0 = 1L * (vote == 0)
  )
head(df)

# Read data
data <- read_poEMirt(
  data = df,
  responses = c("y1", "y0"),
  i = "judge",
  j = "rcid",
  t = "time"
)
summary(data)

# Constraint
con <- which(rownames(data$response) == "Thomas")
con <- rep(con, max(df$time))

# Fit
fit <- poEMirt(
  data = data,
  model = "dynamic",
  constraint = con,
  control = list(
    verbose = 10
  )
)

summary(fit, parameter = "theta")
```

#### Categorical

Using the roll-call vote data of United Nations General Assembly (session 1 to 74) from [`{unvotes}`](https://github.com/dgrtwo/unvotes) by Robinson & Goguen-Compagnoni (2021). Note that the original data is from Voeten (2013). The outcome of interest consists of "yes", "abstain", and "no".

```{r}
data(un_roll_calls, package = "unvotes")
data(un_votes, package = "unvotes")

# Add sessions
df <- un_votes |> 
  left_join(
    un_roll_calls |> 
      select(rcid, session),
    by = "rcid"
  ) 
head(df)

# Create response columns
df <- df |> 
  mutate(tmp = 1) |> 
  spread(
    key = vote, 
    value = tmp
  ) |> 
  mutate(
    across(
      .cols = yes:no,
      .fns = function(x) ifelse(is.na(x), 0, x)
    )
  )
head(df)

# Read data
data <- read_poEMirt(
  data = df,
  responses = c("yes", "abstain", "no"), # "no" is the baseline 
  i = "country",
  j = "rcid",
  t = "session"
)
summary(data)

# Constraint
con <- which(rownames(data$response) == "United States")
con <- rep(con, max(df$session))

# Fit
fit <- poEMirt(
  data = data,
  model = "dynamic",
  constraint = con,
  control = list(
    verbose = 10
  )
)

summary(fit, parameter = "theta")
```

# References

- Goplerud, M. (2019). "A Multinomial Framework for Ideal Point Estimation". *Political Analysis*, 27(1), 69-89.
- Martin A.D., Quinn K.M. & Park J.H. (2011). "MCMCpack: Markov Chain Monte Carlo in R." *Journal of Statistical Software*, 42(9), 22.
- Polson, N. G., Scott, J. G., & Windle, J. (2013). "Bayesian inference for logistic models using Pólya–Gamma latent variables". *Journal of the American statistical Association*, 108(504), 1339-1349.
- Robinson, D., & Goguen-Compagnoni, N. (2021) "unvotes: United Nations General Assembly Voting Data". R package.
- Voeten, E. (2013). *Data and analyses of voting in the United Nations: General Assembly*. Routledge handbook of international organization, 54-66.